Modelling state-switching behaviour in functional data with hidden Markov models

### 2020-09-04 Report/Sections/results.tex
Update report- Evan Edits

Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g. \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For NRKWs, we are interested in categorizing different diving behaviours, with a particular focus on differentiating potential foraging dives. NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKWs for predation \citep{Williams:2009,Noren:2011}. It is therefore of interest to determine the energetic expenditure of NRKWs to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is important to understand the true energetic requirements of a NRKW.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number ###). The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm, when the tagging technology malfunctioned. We smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the vector of observed coarse-scale observations $y = \Big\{y_1,\ldots,y_{267}\Big\}$ is a time series of dive durations in seconds ($s$) with corresponding vector of unobserved dive types $x = \Big\{x_1,\ldots,x_{267}\Big\}$. For dive $t$, the fine-scale observations are contained in $y^*_{t}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$). The corresponding vector of subdive states is $x^*_{t}$. Naturally, $y^* = \Big\{y^*_1,\ldots,y^*_{267}\Big\}$ and $x^* = \Big\{x^*_1,\ldots,x^*_{267}\Big\}$ is the corresponding set of subdive states.

@@

Building a suitable model for these data involves selecting appropriate model structures, emission distributions, and numbers of states on both the coarse and fine scales.

To choose how to model the coarse-scale observations (dive duration), we begin by selecting the model structure. Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations. Visual inspection also shows no obvious intricate structure. Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor transformation is called for.
%
Next, we select the emission distribution of dive duration conditioned on dive type. Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows
a gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
%
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
%
This is consistent with previous studies, including \citet{Barajas:2017}. 
%
Finally, we select the number of dive types for the coarse scale. We do not use information criteria since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}. If the hidden states are well-separated, the lag plot should reveal $N$ distinct patterns where each pattern corresponds to one dive type \citep{Lawler:2019}. This is unfortunately not the case for our killer whale data, but we still consult lag plots together with a heuristic approach based on visual inspection to choose $N = 2$ dive types. The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

To choose how to model the fine-scale observations (acceleration), we again begin with determining model structure. In contrast to the coarse-scale, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $z_t^*$ using Equation (\ref{eqn:z}). We select a window size of two seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
%
We calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to maximum frequencies of 5 Hz for each axis of acceleration. We also down-sample each observation $z^*_t$ such that $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big\}$ to avoid overlapping windows. The resulting sequence of observations is $z^* = \Big\{z^*_1,\ldots,z^*_{267}\Big\}$, where $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},\ldots\Big\}$ and $z^*_{t,t^*} = \{a^*_{t,t^*},w^*_{t,t^*}\}$. We have that $a^*_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $w^*_{t,t^*}$ is a scalar describing the ``wiggliness" of a given window.
%
After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see section 1.1 of the supplementary material). Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in section 2.2. 
%
%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
%
While the $\Ztwo_{t,t^*}$ observations also exhibit some auto-correlation prior to fitting the model, it is less strong than that of $\Zone_{t,t^*}$, and 
%the biological interpretation of auto-correlation between observations of $\Ztwo_{t,t^*}$ is less clear. In addition, 
much of the auto-correlation evident from the lag plot may be explained by subsequent observations having the same subdive state. Therefore, a simple HMM structure is used to model $\Ztwo_{t,t^*}$ for all $t$ and $t^*$. 
%
%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
%

Next, we select the emission distributions for both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$. We assume that the fine-scale emission parameters are shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity. This implies that the subdive states in dive type 1 have the same interpretation as those in dive type 2. 
%
For $\Zone_{t,t^*}$, we assume that each component is Normally distributed with mean and variance defined as in Equation (\ref{eqn:carhmm}). Given that $X^*_{t,t^*} = i^*$, the corresponding unknown parameters are $\phi_A^{*(\cdot,i^*)} \in \mathbb{R}$, $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$. In other words, each axis of acceleration has its own mean and variance, but all axes have the same auto-correlation parameter. We assume that observations of all three axes of acceleration are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
%
On the other hand, we assume that $\Ztwo_{t,t^*}$ is gamma distributed for all $t$ and $t^*$. Further, given that $X^*_{t,t^*} = i^*$, the distribution of $\Ztwo_{t,t^*}$ is parameterized by its mean $\mu_W^{*(\cdot,i^*)} \in \mathbb{R}$ and standard deviation $\sigma_W^{*(\cdot,i^*)} \in \mathbb{R}$. We assume that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.

Finally, to select the number of subdive states $N^*$ we again rely on a heuristic approach based on visual inspection to select $N^* = 3$. Although $N^*$ is selected heuristically, we test the validity of this model using techniques in section \ref{subsec:model_validation}.

The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. On the fine scale, each fine-scale sequence $y^*_t$ is transformed to $z^*$ using Equation (\ref{eqn:z}), $\Zone_{t,t^*}$ is modelled with a CarHMM, and $\Ztwo_{t,t^*}$ is modelled with an HMM for all $t$ and $t^*$. We assume that $\Ztwo_{t,t^*}$ in addition to the components of $\Zone_{t,t^*}$ are all independent of one another and that the subdive types have identical emission distributions across all dive types. In total, the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}.
\end{gather*}
%
Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$.

We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMMs discussed so far. Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model for the CarHHMM-DFT. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. In addition to the CarHHMM-DFT, we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Zone_{t,t^*}$ and $\Ztwo{t,t^*}$ as emissions of a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emssions of a CarHMM. However, the Fourier sums $\Ztwo_{t,t^*}$ are omitted from this model altogether. The CarHHMM therefore does not incorporate the ``wiggliness" of the accelerometer data.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emissions of a CarHMM and $\Ztwo_{t,t^*}$ as emissions an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
%
Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.

@@

To illustrate an application of this method and compare the candidate models, we fit all four models to the data shown in Figure \ref{fig:data}. We first report the results from the full CarHHMM-DFT in detail and then assess the quality of the fit. We then compare the results of the CarHHMM-DFT with those of the other candidate models.

The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}. However, no dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m. It is therefore unlikely that any of the dives used in this study are successful foraging dives.

The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,t^*}$ are separated by an order of magnitude for each subdive state (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,t^*}$ and the smallest variance corresponding to $\Zone_{t,t^*}$. It also has the highest auto-correlation in $\Zone_{t,t^*}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower than in subdive state 1. We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo_{t,t^*}$ and variance of $\Zone_{t,t^*}$ in subdive state 3 are both much higher than in the other two states. The auto-correlation of $\Zone_{t,t^*}$ is also much lower. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions between dives. Within dives, we have
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for dive types 1 and 2.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that about 79\% of dives are short dives of type 1. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 much more often when performing long deep dives compared to short shallow dives (40\% of the time for dive type 1 and 9\% of the time for dive type 2). Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.

>>>>

Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g. \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For NRKWs, we are interested in categorizing different diving behaviours, with a particular focus on differentiating potential foraging dives. NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKWs for predation \citep{Williams:2009,Noren:2011}. It is therefore of interest to determine the energetic expenditure of NRKWs foraging for Chinook as well as the frequency at which NRKWs catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is important to understand the true energetic requirements of a NRKW.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number \#\#\#). The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm due to a malfunction in the tagging technology. To process the data, we smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds ($s$). For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$). Naturally, the collection of all acceleration data is $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.

@@

Defining a suitable model for these data involves selecting an appropriate model structure, emission distributions, and the correct number of hidden states for both the coarse and fine scale observations.

We first select an appropriate model structure for the coarse-scale observations of dive duration. Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection also shows no obvious intricate structure. Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor transformation is called for.
%
Next, we select the emission distribution of dive duration conditioned on dive type. Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows
a gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
%
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
%
This is consistent with previous studies, including that of \citet{Barajas:2017}. 
%
Finally, we select the number of dive types for the coarse scale. We do not use information criteria since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}. If the hidden states are well-separated, the lag plot should reveal $N$ distinct patterns where each pattern corresponds to one dive type \citep{Lawler:2019}. This is unfortunately not the case for our killer whale data, but we still consult lag plots together with a heuristic approach based on visual inspection to choose $N = 2$ dive types. The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).

To model the fine-scale observations of acceleration, we again begin with determining model structure. In contrast to the coarse-scale, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $z_t^*$ using Equation (\ref{eqn:z}). We select a window size of two seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
%
We then calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to maximum frequencies of 5 Hz for each axis of acceleration. We also down-sample each observation $z^*_t$ such that $z^*_t = \left\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\right\}$ and $x^*_t = \left\{x^*_{t,1},x^*_{t,101},x^*_{t,201},\ldots\right\}$ to avoid overlapping windows. The resulting sequence of observations is $z^* = \left\{z^*_1,\ldots,z^*_{267}\right\}$, where $z^*_t = \left\{z^*_{t,1},z^*_{t,101},\ldots\right\}$ and $z^*_{t,t^*} = \left\{a^*_{t,t^*},w^*_{t,t^*}\right\}$. We have that $a^*_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $w^*_{t,t^*}$ is a scalar describing the ``wiggliness" of a given window.
%
After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material). Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in Section 2.2. Due to down-sampling we redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|z^*_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density $\Z_{t,t^*}$ given that $X_t = i$, $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$. We also assume that observations of all three axes of acceleration are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
%
%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
%
While the observations of wiggliness ($\Ztwo_{t,t^*}$) also exhibit some auto-correlation prior to fitting the model, this correlation is less strong than $\Zone_{t,t^*}$, and 
%the biological interpretation of auto-correlation between observations of $\Ztwo_{t,t^*}$ is less clear. In addition, 
much of the auto-correlation evident from the lag plot may be explained by subsequent observations occurring within the same subdive state. Therefore, a simple HMM structure is used to model $\Ztwo_{t,t^*}$ for all $t$ and $t^*$. We assume that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$. 
%
%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
%
%
After defining the model structure, we select the emission distributions for both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$. We assume that the fine-scale emission parameters are shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity. This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2. 
%
For $\Zone_{t,t^*}$, we assume that each component is Normally distributed with mean and variance defined as in Equation (\ref{eqn:carhmm}). Given that $X^*_{t,t^*} = i^*$, the corresponding unknown parameters are $\phi_A^{*(\cdot,i^*)} \in \mathbb{R}$, $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$. In other words, each axis of acceleration has its own mean and variance, but all axes have the same auto-correlation parameter.
%
We assume that $\Ztwo_{t,t^*}$ is gamma distributed for all $t$ and $t^*$. Further, given that $X^*_{t,t^*} = i^*$, the distribution of $\Ztwo_{t,t^*}$ is parameterized by its mean $\mu_W^{*(\cdot,i^*)} \in \mathbb{R}$ and standard deviation $\sigma_W^{*(\cdot,i^*)} \in \mathbb{R}$.
%
Finally, to select the number of subdive states $N^*$ we again rely on a heuristic approach based on visual inspection to select $N^* = 3$. Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.

The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale. Each fine-scale sequence $y^*_t$ is transformed to $z^*$ using Equation (\ref{eqn:z}), $\Zone_{t,t^*}$ is modelled with a CarHMM, and $\Ztwo_{t,t^*}$ is modelled with an HMM for all $t$ and $t^*$. We model $\Zone_{t,t^*}$ to be Normally distributed and $\Ztwo_{t,t^*}$ to be Gamma distributed. We also assume that $\Ztwo_{t,t^*}$ in addition to the components of $\Zone_{t,t^*}$ are all independent of one another and that the subdive types have identical emission distributions across all dive types. In total, the parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \theta = \{\mu^{(1)},\sigma^{(1)},\mu^{(2)},\sigma^{(2)}\} \qquad \text{($Y$ emission parameters), and} \\
    %
    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \theta^{*(\cdot,i^*)} =  \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}.
\end{gather*}
%
Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$.

We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMMs discussed so far. Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model. The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. In addition to the CarHHMM-DFT, we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ as emissions of a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emssions of a CarHMM. However, the Fourier sums $\Ztwo_{t,t^*}$ are omitted from this model altogether. The CarHHMM therefore does not incorporate the ``wiggliness" of the accelerometer data.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emissions of a CarHMM and $\Ztwo_{t,t^*}$ as emissions an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
%
Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.

@@

To illustrate an application of this method and compare the candidate models, we fit all four models to the data shown in Figure \ref{fig:data}. We first report the results from the full CarHHMM-DFT in detail and assess the quality of the fit. We then compare the results of the CarHHMM-DFT with those of the other candidate models.

The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}. However, no dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters. It is therefore unlikely that any of the dives used in this study are successful foraging dives.

The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,t^*}$ are separated by an order of magnitude for each subdive state (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,t^*}$ and the smallest variance corresponding to $\Zone_{t,t^*}$. It also has the highest auto-correlation in $\Zone_{t,t^*}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower than in subdive state 1. We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo_{t,t^*}$ and variance of $\Zone_{t,t^*}$ in subdive state 3 are both much higher than in the other two states. The auto-correlation of $\Zone_{t,t^*}$ is also much lower. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions between dives. Within dives, we have
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for dive types 1 and 2.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that about 79\% of dives are short dives of type 1. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 40\% of the time when in a dive of type 1 but only 9\% of the time when in a dive of type 2. Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.
